Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-02-25 paper/main.tex
reframed some ideas in discussion to clarify 'sweet spot' idea

% supplementary tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}
% supplementary figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\forcesCorrs}{3}
\newcommand{\bosCorrs}{4}
\newcommand{\jaccard}{5}
\newcommand{\ldaVsBERT}{6}
\newcommand{\individualKnowledgeMapsA}{7}
\newcommand{\individualKnowledgeMapsB}{8}
\newcommand{\individualKnowledgeMapsC}{9}
\newcommand{\individualLearningMapsA}{10}
\newcommand{\individualLearningMapsB}{11}
% supplementary discussion
\newcommand{\suppResults}{\textit{Supplementary results}}

@@

Our work makes several contributions to the study of how people acquire
conceptual knowledge. First, from a methodological standpoint, our modeling
framework provides a systematic means of mapping out and characterizing
knowledge in maps that have infinite (arbitrarily many) numbers of coordinates,
and of ``filling out'' those maps using relatively small numbers of 
multiple-choice quiz questions. Our experimental finding that we can use these 
maps to predict responses to held-out questions has several psychological 
implications as well. For example, concepts that are assigned to nearby 
coordinates by the text embedding model also appear to be ``known to a similar 
extent'' (as reflected by participants' responses to held-out questions;
Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge ``falls off'' with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

@@

At the opposite end of the spectrum from large language models, one could also
imagine using an even \textit{simpler} ``model'' than LDA that relates the 
contents of course lectures and quiz questions through explicit word-overlap 
metrics (rather than similarities in the latent topics they exhibit). In a 
supplementary analysis (Supp.~Fig.~\jaccard), we compared the LDA-based 
question-lecture matches shown in Figure~\ref{fig:question-correlations} with
analogous matches based on the Jaccard similarity between each question's text 
and each sliding window from the corresponding lecture's transcript. Similarly 
to the embeddings derived from BERT, we found that this approach also blurred 
meaningful distinctions between concepts presented in different parts of each 
lecture and tested by different quiz questions. In this case, this lack of 
specificity reflects that meaningful distinctions between concepts do not tend
to emerge at the level of individual words.

In this way, we view our approach as occupying a sort of ``sweet spot'' between
much larger language models and simpler word-matching-based approaches, in that 
it enables us to capture the relevant conceptual content of course materials at 
the appropriate semantic scale. Our approach enables us to accurately and 
consistently identify each question's content in a way that matches it with 
specific content from the lectures and distinguishes it from other questions 
about similar content. In turn, this enables us to construct accurate 
predictions about participants' knowledge of the conceptual content tested by 
individual quiz questions (Fig.~\ref{fig:predictions}).

Another application for large language models that does \textit{not} require
explicitly modeling the content of individual lectures or questions is to
leverage the models' abilities to generate text. For example, generative text
models like ChatGPT~\citep{ChatGPT} and LLaMa~\citep{TouvEtal23} are already
being used to build a new generation of interactive tutoring
systems~\citep[e.g.,][]{MannEtal23b}. Unlike the approach we have taken here,
these generative text model-based systems do not explicitly model what learners
know, or how their knowledge changes over time with training. One could imagine
building a hybrid system that combines the best of both worlds: a large
language model that can \textit{generate} text, combined with a smaller model
that can \textit{infer} what learners know and how their knowledge changes over
time. Such a hybrid system could potentially be used to build the next
generation of interactive tutoring systems that are able to adapt to learners'
needs in real time, and that are able to provide more nuanced feedback about
what learners know and what they do not know.

>>>>

% supplementary tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}
% supplementary figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\forcesCorrs}{3}
\newcommand{\bosCorrs}{4}
\newcommand{\jaccard}{5}
\newcommand{\ldaVsBERT}{6}
\newcommand{\individualKnowledgeMapsA}{7}
\newcommand{\individualKnowledgeMapsB}{8}
\newcommand{\individualKnowledgeMapsC}{9}
\newcommand{\individualLearningMapsA}{10}
\newcommand{\individualLearningMapsB}{11}
% supplementary results
\newcommand{\suppResults}{\textit{Supplementary results}}

@@

Our work makes several contributions to the study of how people acquire
conceptual knowledge. First, from a methodological standpoint, our modeling
framework provides a systematic means of mapping out and characterizing
knowledge in maps that have infinite (arbitrarily many) numbers of coordinates,
and of ``filling out'' those maps using relatively small numbers of 
multiple-choice quiz questions. Our experimental finding that we can use these 
maps to predict success on held-out questions has several psychological 
implications as well. For example, concepts that are assigned to nearby 
coordinates by the text embedding model also appear to be ``known to a similar 
extent'' (as reflected by participants' responses to held-out questions;
Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge ``falls off'' with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

@@

At the opposite end of the spectrum from large language models, one could also
imagine using an even \textit{simpler} ``model'' than LDA that relates the 
contents of course lectures and quiz questions through explicit word-overlap 
metrics (rather than similarities in the latent topics they exhibit). In a 
supplementary analysis (Supp.~Fig.~\jaccard), we compared the LDA-based 
question-lecture matches shown in Figure~\ref{fig:question-correlations} with
analogous matches based on the Jaccard similarity between each question's text 
and each sliding window from the corresponding lecture's transcript. Similarly 
to the embeddings derived from BERT, we found that this approach also blurred 
meaningful distinctions between concepts presented in different parts of each 
lecture and tested by different quiz questions. But rather than characterizing
their contents at too \textit{broad} a semantic scale, the lack of specificity 
in this approach arises from considering too \textit{narrow} a semantic scale: 
the sorts of concepts typically conveyed in course lectures and tested by quiz 
questions are not defined (and meaningful similarities and distinctions between 
them do not tend to emerge) at the level of individual words.

In other words, while the embedding spaces of more complex large language models 
afford low resolution at the scale of individual course lectures and questions 
because they ``zoom out'' too far, simpler word-matching measures yield low 
resolution because they ``zoom \textit{in}'' too far. In this way, we view our 
approach as occupying a sort of ``sweet spot'' between simpler and more complex 
alternatives, in that it enables us to characterize the contents of course 
materials at the appropriate semantic scale where relevant concepts ``come into 
focus.'' Our approach enables us to accurately and consistently identify each 
question's content in a way that matches it with specific content from the 
lectures and distinguishes it from other questions about similar content. In 
turn, this enables us to construct accurate predictions about participants' 
knowledge of the conceptual content tested by individual quiz questions 
(Fig.~\ref{fig:predictions}).

Another application for large language models that does \textit{not} require
explicitly modeling the content of individual lectures or questions is to
leverage these models' abilities to generate text. For example, generative text
models like ChatGPT~\citep{ChatGPT} and LLaMa~\citep{TouvEtal23} are already
being used to build a new generation of interactive tutoring
systems~\citep[e.g.,][]{MannEtal23b}. Unlike the approach we have taken here,
these generative text model-based systems do not explicitly model what learners
know, or how their knowledge changes over time with training. One could imagine
building a hybrid system that combines the best of both worlds: a large
language model that can \textit{generate} text, combined with a smaller model
that can \textit{infer} what learners know and how their knowledge changes over
time. Such a hybrid system could potentially be used to build the next
generation of interactive tutoring systems that are able to adapt to learners'
needs in real time, and that are able to provide more nuanced feedback about
what learners know and what they do not know.
